### 2014-05-28 patchy-selection-paper.tex
missing rho
haplotype within a patch section rewrite

%%%%%%% %%%%%%%%%%
\subsection{Length of the shared haplotype within a patch}

To really make use of the expressions in the previous section for the length of a haplotype shared
between patches that are coadapted through migration,
we need to know the length of the haplotype shared by individuals \emph{within} a patch.
A detailed treatment of this is well beyond the scope of this paper,
but we can get a good idea of the scale through heuristic arguments.
To do this, we compare the rates of coalescence to the rate of recombination between B and b haplotypes.

Sharing is ultimately determined by the coalescence time within a patch at the B allele.
We could make arguments about coalescence between lineages moving in the patch \citep[as in][]{hallatschek},
but given the well-known difficulties to spatial coalescence \citep{felsenstein1978pain,bartonetheridgedepaulis},
it seems reasonable to treat the patch as panmictic as far as coalescence is concerned.
In other words, we suppose that all adapted individuals in a given patch have inherited their B allele from a single common ancestor living in the patch,
and that the time back to the common ancestor at the selected locus between two randomly sampled individuals
is exponentially distributed, with mean inversely proportional to the patch size, $1/(A\rho)$.
If this is sufficiently shorter than the typical time to a common ancestor with a b allele,
then the signal of a shared haplotype will be noticible.

Consider a site at some distance to the selected locus, in two chromosomes carrying the B allele.
The most recent common ancestor at this site may not be the same as that at the selected locus,
but the two chromosomes will still share a haplotype at this site
if coalescence at this site occurs before recombination onto the b background.
In other words, if all ancestral chromosomes along the lineages connecting the two chromosomes to their most recent common ancestor
carried the B allele,
they share a B haplotype at that site.
Therefore, we need to understand how often the ancestral lineage at a site linked to the selected locus
moves between b and B backgrounds, as it moves across geography.
\plr{alternate setup:}
Consider the genotypes of two B-carrying chromosomes at a site linked to the selected site.
The main thing we need to know is which of the following lived more recently:
their most recent common ancestor at that site,
or an ancestor of either of them at that site that carried the b allele.

First consider the lineage of a single B allele, moving back through time.
Suppose that the B allele is at mutation-selection balance, 
with the local proportion of B alleles near $x$ equal to $p(x)$,
and that we are close enough to the patch that fluctuations are small (unlike in previous sections).
If we suppose that type B individuals at location $y$ reproduce at rate $1+s(y)$,
and the chance their offspring move to $x$ is $m(y,x)$ (with $\sum_x m(y,x)=1$),
then the total rate of influx of B alleles from $y$ into $x$ is $\rho p(y) (1+s(y)) m(y,x)$.
As argued by \citet{hudson},
in a discrete model the rate at which a B lineage moves from $x$ to $y$
is this influx divided by the local abundance of $B$ alleles:
\begin{align}
  r(x,y) = \frac{ p(y) (1+s(y)) m(y,x) }{ p(x) } .
\end{align}
After the same large-population, continuous-space limit that underlies the rest of this paper,
this motion should converge to a diffusion process \citep[that is similar to that studied by][]{hallatschek},
notwithstanding some technical difficulties (e.g.\ discontinuity of $s(x)$).
In It\^o differential notation, 
the location $X_t$ of a B lineage $t$ generations ago satisfies
\begin{align}
  d X_t &= (1+s(X_t)) D \log( p(X_t) (1+s(X_t)) ) dt + \sigma \sqrt{1+s(X_t)} dB_t,\\
    % &=  ( \log(p(X_t))' (1+s(X_t)) + s'(X_t) ) dt + \sqrt{1+s(X_t)} dB_t,\\
\end{align}
where $s(x)=s_p$ if $x$ is in the patch and $s(x)=s_m$ otherwise, 
$B_t$ is a standard Brownian motion, and $D$ denotes the gradient operator.
The main effect is the drift term, that pulls the lineage towards regions where $p(x)$ is larger;
since $s$ is generally small, the speed of the Brownian noise term has little effect.
Furthermore, it is easy to verify that the stationary distribution of this process is
\begin{align}
  \pi(x) = \frac{ p(x)^2 (1+s(x)) }{ \int_y p(y)^2 (1+s(y)) dy } .
\end{align}

If we are instead following a lineage at a locus at genetic distance $r$ from the selected locus,
and begin linked to a B allele,
then the lineage will move as a B lineage up until the most recent time
it is found linked to a b allele.
If the lineage is at $x$, the probability that a recombination that moves it from a B background
to a b background occurs in time $dt$ is $dt (1-p(x)) (1+s(x)/2)$.
The Feynman-Kac formula could then be used to write down differential equations
solved by various moments of the haplotype length as a function of the sampling location of the lineage.

More simply, consider: within the patch, where $p(x) \approx 1$, 
a linked lineage moves roughly as Brownian motion, and is unlikely to recombine onto a b background.
The rate of recombination between B and b backgrounds depends on the proportion of the patch in which $p(x)$ is intermediate.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
(and, therefore, the rate of recombination)
is roughly $r (\sigma/\sqrt{s})A^{-1/d}$.
This will be equal to the coalescence rate if
$1/(A\rho) = r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the length of linkage shared within the patch is on the order of
\begin{align}
  \frac{1}{r} = \frac{ \sigma A^{(d-1)/d} }{\sqrt{s}} ,
\end{align}
which does not increase with $A$ in one dimension,
but increases with $\sqrt{A}$ in two dimensions.

>>>>

%%%%%%% %%%%%%%%%%
\subsection{Length of the shared haplotype within a patch}
\label{ss:patch_haplotype}

To really make use of the expressions in the previous section,
we need to know the length of the haplotype shared by individuals \emph{within} a patch.
A detailed treatment of this is well beyond the scope of this paper,
but we can get a good idea of the scale through heuristic arguments.
To do this, we compare the rate of recombination between B and b haplotypes
to the rate of coalescence of B haplotypes within a patch,
which we take to be $1/(A\rho)$ 
(bypassing the well-known difficulties to spatial coalescence \citep{felsenstein1978pain,bartonetheridgedepaulis}).

Consider the lineage of a single B allele, moving back through time,
in a patch at mutation-selection balance, 
i.e.\ with the local proportion of B alleles near $x$ equal to $p(x)$,
and that the lineage stays close enough to the patch 
that fluctuations about $p(x)$ are small (unlike in previous sections).
If we suppose that type B individuals at location $y$ reproduce at rate $1+s(y)$,
and the chance their offspring move to $x$ is $m(y,x)$ (with $\sum_x m(y,x)=1$),
then the total rate of influx of B alleles from $y$ into $x$ is $\rho p(y) (1+s(y)) m(y,x)$.
As argued by \citet{hudson}, the rate at which a B lineage moves from $x$ to $y$
is this influx divided by the local abundance of $B$ alleles,
or $\frac{ p(y) (1+s(y)) m(y,x) }{ p(x) }$.
After the same large-population, continuous-space limit that underlies the rest of this paper,
this motion should converge to a diffusion process \citep[as assumed by][]{hallatschek},
notwithstanding some technical difficulties (e.g.\ discontinuity of $s(x)$).
In It\^o differential notation, 
the location $X_t$ of a B lineage $t$ generations ago satisfies
\begin{align}
    d X_t &= (1+s(X_t)) \frac{d}{d\,x} \log( p(X_t) (1+s(X_t)) ) dt + \sigma \sqrt{1+s(X_t)} dB_t,\\
    % &=  ( \log(p(X_t))' (1+s(X_t)) + s'(X_t) ) dt + \sqrt{1+s(X_t)} dB_t,\\
\end{align}
where $s(x)=s_p$ if $x$ is in the patch and $s(x)=s_m$ otherwise, 
and $B_t$ is a standard Brownian motion.
This is a diffusion that, when at $x$, 
is pulled toward regions with more B alleles
by the drift term $(1+s(x))\frac{d}{d\,x} \log( p(x) (1+s(x)) )$,
and moves at nearly constant speed (since $\sqrt{1+s(x)}$ is close to 1).
Furthermore, it is easy to verify that the stationary distribution of this process is
\begin{align}
  \pi(x) = \frac{ p(x)^2 (1+s(x)) }{ \int_y p(y)^2 (1+s(y)) dy } .
\end{align}

A lineage at a locus at genetic distance $r$ that begins linked to a B allele
moves as a B lineage up until the most recent time
it recombined from a b background to a B background,
which occurs when at $x$ with rate $r (1-p(x)) (1+s(x)/2)$.
The Feynman-Kac formula could then be used to write down differential equations
solved by various moments of the haplotype length as a function of the sampling location of the lineage.

Or, more simply, if recombination is on a slower time scale than mixing within the patch,
then the time until the last recombination is the integral of $r (1-p(x)) (1+s(x)/2)$ 
against the stationary distribution.
Within the patch, $p(x) \approx 1$,
so recombination between backgrounds is unlikely to occur,
and so the rate of recombination depends to a first approximation 
on the proportion of the patch in which $p(x)$ is intermediate.
This region where $p(x)$ is intermediate has width of order $\sigma/\sqrt{s}$, where $s = \min(s_p,s_m)$.
As we increase the size of a patch, the area on which $p(x)$ is intermediate scales with the perimeter of the patch.
Therefore, the proportion of the time a lineage at stationarity is found in the intermediate regions
is roughly $2(\sigma/\sqrt{s})A^{-1/d}$,
and so the rate of recombination of a linked lineage to the b background
is, taking $p(x) \approx 1/2$ in this region, roughly $r (\sigma/\sqrt{s})A^{-1/d}$,
This will be equal to the coalescence rate if
$1/(A\rho) = r (\sigma/\sqrt{s}) A^{-1/d}$, 
and hence the length of linkage shared within the patch is on the order of
\begin{align}
  \frac{1}{r} = \frac{ \sigma \rho A^{(d-1)/d} }{\sqrt{s}} ,
\end{align}
which does not increase with $A$ in one dimension,
but increases with $\sqrt{A}$ in two dimensions.
